Weather covariance matrices as U_hyp for GR50

### 2021-01-13 R/Analysis_v0.6_weather_covariance_Uhyp_greenup.Rmd
Modify nohup mash scripts to remove self-correlation of phenotypes and add greenup corr for flowering

### run mash 
```{r run mash with gulf}
for(k in c(1,3,2)){

  numSNPs <- 19000
  m_out <- mash_standard_run(path = file.path(outputdir), #, names(subpop_v)[k]), # use to run on genegenie within mash dirs
                             numSNPs = numSNPs, 
                             U_hyp = U_hyp,
                             suffix = paste0("GR50_", names(subpop_v)[k]),
                             saveoutput = TRUE)
}

inputfiles
```

@@

phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -41) %>%
    str_replace(., "_$", "")
phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
out1phe <- pvdiv_bigsnp2mashr(path = file.path(outputdir), 
                           snp = snp, gwas_rds = gwas_rds, 
                           phenotypes = phenotypes, numSNPs = numSNPs, 
                           clump = TRUE, scaled = TRUE, saveoutput = TRUE,
                           suffix = paste0("GR50_", names(subpop_v)[k]))
m_out <- mash_standard_run(path = file.path(outputdir), #, names(subpop_v)[k]), # use to run on genegenie within mash dirs
                             numSNPs = numSNPs, 
                             suffix = paste0("GR50_", names(subpop_v)[k]),
                             saveoutput = TRUE)
saveRDS(m_out, file = file.path(mashdir, paste0("Strong_Effects_", numSNPs,
                                                "GR50_", names(subpop_v)[k],
                                                "_no_Uhyp.rds")))
m_out1 <- mash_standard_run(path = file.path(outputdir), numSNPs = numSNPs, 
                           U_ed = file.path(outputdir, paste0("Model_of_",
                           "random_tests_24000SNPs_GR50_Midwest.rds")),
                           U_hyp = U_hyp,
                           suffix = paste0("GR50_", names(subpop_v)[k]),
                           saveoutput = FALSE)
saveRDS(m_out1, file = file.path(mashdir, paste0("Strong_Effects_", numSNPs,
                                                "GR50_", names(subpop_v)[k],
                                                "_no_Uhyp.rds")))
```

@@

## 3b. Covar Loadings
How many SNP effects load onto each covariance matrix?
```{r load mash objects}
gulfm <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf/Strong_Effects19000SNPs_GR50_Gulf_Uhyp.rds")

gandm <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf_and_Midwest/Strong_Effects19000SNPs_GR50_Gulf_and_Midwest_Uhyp.rds")
```
44765 SNPs tested for G&M, 45931 SNPs tested for Gulf.

Midwest I haven't run successfully yet as of 2021-01-12 - I will need to remove some locations with all-zero effects, probably.

@@

covargulf <- mash_plot_covar(gulfm, suffix = "19000SNPs_GR50_Gulf", saveoutput = TRUE)
covargulf
get_U_by_mass(gulfm)
mash_plot_Ulist(gulfm, range = get_U_by_mass(gulfm), limits = FALSE)
mash_plot_manhattan_by_condition(gulfm, saveoutput = FALSE)
mash_plot_sig_by_condition(gulfm, saveoutput = FALSE)
```

@@

```{r}
covargam$covar_df %>% 
  mutate(Type = case_when(grepl("ED_", `Covariance Matrix`) ~ "data-driven",
                          grepl("single_effect", `Covariance Matrix`) ~ "canonical",
                          grepl("simple_het", `Covariance Matrix`) ~ "canonical",
                          grepl("no_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("identity", `Covariance Matrix`) ~ "canonical",
                          grepl("equal_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("_and_Midwest$", `Covariance Matrix`) ~ "GAM_hyp",
                          grepl("Midwest$", `Covariance Matrix`) ~ "Midw_hyp",
                          grepl("Gulf$", `Covariance Matrix`) ~ "Gulf_hyp",
                          TRUE ~ "other"
                          )) %>%
  group_by(Type) %>%
  summarise(Mass = sum(Mass))
```
Gulf: 20% canonical, 68.5% data-driven, 9.5% Gulf hyp, 1.6% Midwest hyp
Gulf & Midwest: 28% canonical, 62% data-driven, 5.5% Gulf, 1.2% Midwest, 2.8% G&M

>>>>

### run mash

Ran mash using six 'nohup_v0.6' R scripts on the server 2021-01-12.

@@

phenotypes_1 <- str_sub(gwas_rds, start = 16, end = -41) %>%
    str_replace(., "_$", "")
phenotypes <- paste0(names(subpop_v)[k], "_", phenotypes_1)
numSNPs <- round(ceiling(1200000/length(phenotypes)^2)/1000)*1000
out1phe <- pvdiv_bigsnp2mashr(path = file.path(outputdir), 
                           snp = snp, gwas_rds = gwas_rds, 
                           phenotypes = phenotypes, numSNPs = numSNPs, 
                           clump = TRUE, scaled = TRUE, saveoutput = TRUE,
                           suffix = paste0("GR50_", names(subpop_v)[k]))
# Ran mash using a 'nohup_v0.6' R script.
```

@@

## 3a. Log-likelihoods
```{r}
m_gulf_gr_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf/Strong_Effects_19000_SNPs_GR50_Gulf_Uhyp.rds")
m_gulf_gr_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf/Strong_Effects_19000_SNPs_GR50_Gulf_no_Uhyp.rds")
m_gulf_fl_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf/Strong_Effects_19000_SNPs_FL50_Gulf_Uhyp.rds")
m_gulf_fl_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf/Strong_Effects_19000_SNPs_FL50_Gulf_no_Uhyp.rds")

m_midw_gr_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Midwest/Strong_Effects_24000_SNPs_GR50_Midwest_Uhyp.rds")
m_midw_gr_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Midwest/Strong_Effects_24000_SNPs_GR50_Midwest_no_Uhyp.rds")  
m_midw_fl_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Midwest/Strong_Effects_33000_SNPs_FL50_Midwest_Uhyp.rds")
m_midw_fl_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Midwest/Strong_Effects_33000_SNPs_FL50_Midwest_no_Uhyp.rds")

m_gam_gr_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf_and_Midwest/Strong_Effects_19000_SNPs_GR50_Gulf_and_Midwest_Uhyp.rds")
m_gam_gr_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf_and_Midwest/Strong_Effects_19000_SNPs_GR50_Gulf_and_Midwest_no_Uhyp.rds")
m_gam_fl_hyp <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf_and_Midwest/Strong_Effects_19000_SNPs_FL50_Gulf_and_Midwest_Uhyp.rds")
m_gam_fl_no <- readRDS("~/Github/pvdiv-phenology-gxe/analysis/mash/Gulf_and_Midwest/Strong_Effects_19000_SNPs_FL50_Gulf_and_Midwest_no_Uhyp.rds")
```

```{r}
m_gulf_gr_hyp$loglik
m_gulf_gr_no$loglik
m_gulf_fl_hyp$loglik
m_gulf_fl_no$loglik

-2*(m_gulf_gr_no$loglik - m_gulf_gr_hyp$loglik) # negative
-2*(m_gulf_fl_no$loglik - m_gulf_fl_hyp$loglik) # positive
# test statistic equal to twice the absolute difference in these log likelihoods is assumed to be distributed as Chi square with one degree of freedom. Larger positive statistics are more significant.
1 - pchisq(-2*(m_gulf_fl_no$loglik - m_gulf_fl_hyp$loglik), 30) #sig
1 - pchisq(-2*(m_gulf_gr_no$loglik - m_gulf_gr_hyp$loglik), 15)
# I don't know how many df each covariance matrix (each Uhyp) would take. At least one per added Uhyp? Up to (ncond*ncond)/2 per added Uhyp? What do they do in the mash paper?

-2*(m_midw_gr_no$loglik - m_midw_gr_hyp$loglik) # positive
-2*(m_midw_fl_no$loglik - m_midw_fl_hyp$loglik) # v. negative 
1 - pchisq(-2*(m_midw_fl_no$loglik - m_midw_fl_hyp$loglik), 30*32)
1 - pchisq(-2*(m_midw_gr_no$loglik - m_midw_gr_hyp$loglik), 15*32) # sig

-2*(m_gam_gr_no$loglik - m_gam_gr_hyp$loglik)  # positive
-2*(m_gam_fl_no$loglik - m_gam_fl_hyp$loglik) # v. positive
1 - pchisq(-2*(m_gam_fl_no$loglik - m_gam_fl_hyp$loglik), 30*32) # sig
1 - pchisq(-2*(m_gam_gr_no$loglik - m_gam_gr_hyp$loglik), 15*32) # sig

1-pchisq(0.0279,1) # LR -2(L0 - L1) gives LR of 0.0279 which gives p of 0.87
1-pchisq((6.89), 1) # LR of 6.89 gives p of ~0.01
```

A test using the log-likelihood

For the MLR estimator there is an additional test for nested models. This test compares the log-likelihoods for the null and alternative models rather than the chi-square values. Below is a list of the information needed, along with the symbol (i.e., letter and number) used to represent each value.

L0 = log-likelihood for the null model.
L1 = log-likelihood for the alternative model.

c0 = scaling correction factor for the null model.
c1 = scaling correction factor for the alternative model.

p0 = number of parameters estimated in the null model.
p1 = number of parameters estimated in the alternative model.

From this information the tests statistic TRd can be calculated (note that cd is calculated first then used to calculate TRd), along with the degrees of freedom:

cd = (p0*c0-p1*c1)/(p0-p1)
TRd = -2*(L0-L1)/cd
df = p1-p0

## 3b. Covar Loadings
How many SNP effects load onto each covariance matrix?

```{r}
mash_plot_covar(m_gulf_gr_no)
mash_plot_covar(m_gulf_gr_hyp)
mash_plot_covar(m_gulf_fl_no)
mash_plot_covar(m_gulf_fl_hyp)

mash_plot_covar(m_midw_gr_no)
mash_plot_covar(m_midw_gr_hyp)
mash_plot_covar(m_midw_fl_no)
mash_plot_covar(m_midw_fl_hyp)

mash_plot_covar(m_gam_gr_no)
mash_plot_covar(m_gam_gr_hyp)
mash_plot_covar(m_gam_fl_no)
mash_plot_covar(m_gam_fl_hyp)
```

```{r}
cov_gg <- mash_plot_covar(m_gulf_gr_hyp)
cov_gf <- mash_plot_covar(m_gulf_fl_hyp)

cov_mg <- mash_plot_covar(m_midw_gr_hyp)
cov_mf <- mash_plot_covar(m_midw_fl_hyp)

cov_gamg <- mash_plot_covar(m_gam_gr_hyp)
cov_gamf <- mash_plot_covar(m_gam_fl_hyp)

cov_gg$ggobject + ggtitle("Gulf greenup 11.4%")
cov_gf$ggobject + ggtitle("Gulf flowering 32.6%")
cov_mg$ggobject + ggtitle("Midwest greenup 35.6%")
cov_mf$ggobject + ggtitle("Midwest flowering 45.7%")
cov_gamg$ggobject + ggtitle("Gulf and Midwest greenup 9.6%")
cov_gamf$ggobject + ggtitle("Gulf and Midwest flowering 46%")
```

46% of effect mass for both subpop flowering was on hyp matrices, while only 9.6% of greenup was.
Gulf greenup: 11.4%  flowering: 32.6%
Midwest greenup: 35.6% flowering 45.7%

```{r}
cov_gamf$covar_df %>% 
  mutate(Type = case_when(grepl("ED_", `Covariance Matrix`) ~ "data-driven",
                          grepl("single_effect", `Covariance Matrix`) ~ "canonical",
                          grepl("simple_het", `Covariance Matrix`) ~ "canonical",
                          grepl("no_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("identity", `Covariance Matrix`) ~ "canonical",
                          grepl("equal_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("_and_Midwest$", `Covariance Matrix`) ~ "GAM_hyp",
                          grepl("Midwest$", `Covariance Matrix`) ~ "Midw_hyp",
                          grepl("Gulf$", `Covariance Matrix`) ~ "Gulf_hyp",
                          TRUE ~ "other"
                          )) %>%
  group_by(Type) %>%
  summarise(Mass = sum(Mass))

cov_gamg$covar_df %>% 
  mutate(Type = case_when(grepl("ED_", `Covariance Matrix`) ~ "data-driven",
                          grepl("single_effect", `Covariance Matrix`) ~ "canonical",
                          grepl("simple_het", `Covariance Matrix`) ~ "canonical",
                          grepl("no_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("identity", `Covariance Matrix`) ~ "canonical",
                          grepl("equal_effects", `Covariance Matrix`) ~ "canonical",
                          grepl("_and_Midwest$", `Covariance Matrix`) ~ "hyp",
                          grepl("Midwest$", `Covariance Matrix`) ~ "hyp",
                          grepl("Gulf$", `Covariance Matrix`) ~ "hyp",
                          TRUE ~ "other"
                          )) %>%
  group_by(Type) %>%
  summarise(Mass = sum(Mass))
```
Gulf: 20% canonical, 68.5% data-driven, 9.5% Gulf hyp, 1.6% Midwest hyp
Gulf & Midwest: 28% canonical, 62% data-driven, 5.5% Gulf, 1.2% Midwest, 2.8% G&M

46% of effect mass for both subpop flowering was on hyp matrices, while only 9.6% of greenup was.
Gulf greenup: 11.4%  flowering: 32.6%
Midwest greenup: 35.6% flowering 45.7%

44765 SNPs tested for G&M, 45931 SNPs tested for Gulf.

@@

covargulf <- mash_plot_covar(m_gulf_fl_hyp, suffix = "19000SNPs_GR50_Gulf", saveoutput = FALSE)
covargulf
get_U_by_mass(gulfm)
mash_plot_Ulist(gulfm, range = get_U_by_mass(gulfm), limits = FALSE)
mash_plot_manhattan_by_condition(gulfm, saveoutput = FALSE)
mash_plot_sig_by_condition(gulfm, saveoutput = FALSE)
```

@@
